---
title: This is a sample .qmd post
author: Andrew
date: "2023-07-01"
slug: sample-post
summary: |
  Working on understanding how to mingle content from Quarto into sveltekit.
draft: true
knitr:
  opts_knit: 
    base.url: "/"
  opts_chunk:
    fig.path: "./+page_files/"
    class-output: qmdresults
    class-message: qmdmessage
format:
  gfm:
    variant: +yaml_metadata_block
---

<script>
  import Summary from "./metadata_getter.svelte";
  import Head from "$lib/components/Head.svelte";
</script>

<Head {title} />

# {title}
**{date}**

Hello. Today's date is {date}.

<!-- This is just an example of how you can easily bring in components! -->
<Summary summary="{summary}" />

What have I figured out about using Quarto/Mdsvex?:

1. `fig.path` needs to be set to `"./+page_files/"
    - you can then use the relativeImages npm package to properly link to output files

2. If you add **both** `fig.cap` and `fig.alt`, plot outputs will be wrapped with a `<figure>` tag. This could be helpful? Maybe the stuff can be styled now...

3. Adding `class-output` and `class-message` to `opts_chunk` wraps R code chunk output/messages with "``` qmdoutput". Not sure how this is then parsed by Mdsvex, but it's something?

4. Layouts ensure you only have to load CSS once?

## Todo

- [x] plot output (relative image paths)
- [x] how to do math
- [x] code highlighting
    - gonna say this is done, 90% of what I've been looking for
- [ ] asides?
- [x] footnotes?
    - had to run `npm add --save-dev remark-footnotes@2.0` in order for this to work
- [ ] citations?
- [x] table styling
    - think I'm happy with this for now, will probably need tuning
    - using `format = "html"` seems to be really useful for `knitr` tables
- [x] ggplot theme
    - where to store? `assets/`?

## Styles?

Can we add styles?

Hell yes we can.

## Testing output/rendering

Here is a quote:

> The asphault city is my home. Furnished
> 
> From the outset with every sacramental prerequisite:
> 
> With newspapers. And tobacco. And brandy.
> 
> Distrustful and idle and contented to the end.

### Math

Here is some math.[^1]

$$
\int_0^1 x^3 + \theta x^2\ dx
$$

::: {.column-margin}
We know from *the first fundamental theorem of calculus* that for $x$ in $[a, b]$:

$\frac{d}{dx}\left( \int_{a}^{x} f(u)\ du\right)=f(x)$.
:::

Note that Math can also be inline, like this: $y = \beta_0 + \beta_1 x_1 + \epsilon$.

### Code output

```{r}
library(tidyverse)

source("../../../../lib/assets/theme-ndrewwm.R")
```

```{r}
summary(airquality)
```

Some more output.^[Pay this no mind.]

```{r}
lm(mpg ~ wt + cyl, data = mtcars) |> summary()
```

### Tables

Here's a simple table.

```{r}
mtcars |>
  group_by(vs) |>
  summarize(m = mean(mpg), s = sd(mpg)) |>
  knitr::kable(digits = 2)
```

Here are a few that are larger.

```{r}
library(palmerpenguins)

fit <- lm(flipper_length_mm ~ bill_length_mm + body_mass_g + sex + species, data = penguins)

fit |>
  broom::tidy() |>
  knitr::kable(
    digits = 2,
    caption = "These are regression coefficients, and associated standard errors.",
    format = "html"
  )

fit |> broom::glance() |> knitr::kable(digits = 2)
```

### Plots

Plots are messy though.^[Hey, you can also create in-line footnotes. They seem super convenient.]

```{r, dev = "png", dev.args=list(bg="transparent")}
#| fig-align: center
#| fig-cap: "This is the caption."
#| fig-alt: "Maybe some alt-text?"

theme_set(theme_ndrewwm())
 
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) + geom_boxplot()
```

Here's another plot.

```{r, dev = "png", dev.args=list(bg="transparent")}
#| fig-cap: "This is the caption."
#| fig-alt: "Maybe some alt-text?"

library(palmerpenguins)

ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point()

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_density()

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram()

ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram() +
  facet_wrap(~island)
```

Sick, this is working now.[^2]

[^1]: There might be footnotes.

[^2]: You could consider writing something that's incredibly long as a footnote.

    These long footnotes could even span multiple paragraphs. However, note that new paragraphs need to be indented.
