---
title: This is my title.
knitr:
  opts_knit: 
    base.url: "/"
  opts_chunk:
    fig.path: "./+page_files/"
    class-output: qmdresults
    class-message: qmdmessage
format: gfm
---

Hello. What have I figured out about using Quarto/Mdsvex?:

1. `fig.path` needs to be set to `"./+page_files/"
  - you can then use the relativeImages npm package to properly link to output files

2. If you add **both** `fig.cap` and `fig.alt`, plot outputs will be wrapped with a `<figure>` tag. This could be helpful? Maybe the stuff can be styled now...

3. Adding `class-output` and `class-message` to `opts_chunk` wraps R code chunk output/messages with "``` qmdoutput". Not sure how this is then parsed by Mdsvex, but it's something?

## Todo

- [x] plot output (relative image paths)
- [ ] how to do math
- [ ] code highlighting

## Styles?

Can we add styles?

<style>
  h2 {
    color: orange;
  }
</style>

Hell yes we can.

## Testing output/rendering

```{r}
library(tidyverse)
```

```{r}
summary(airquality)
```

Some more output.

```{r}
lm(mpg ~ wt + cyl, data = mtcars) |> summary()
```

```{r}
mtcars |>
  group_by(vs) |>
  summarize(m = mean(mpg), s = sd(mpg)) |>
  knitr::kable(digits = 2)
```

Plots are messy though.

```{r}
#| fig-align: center
#| fig-cap: "This is the caption."
#| fig-alt: "Maybe some alt-text?"
 
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) + geom_boxplot()
```

Sick, this is working now.
