---
title: This is where the title would go, if that's important.
author: Andrew
date: 7-3-2023
summary: |
  Working on understanding how to mingle content from Quarto into sveltekit.
knitr:
  opts_knit: 
    base.url: "/"
  opts_chunk:
    fig.path: "./+page_files/"
    class-output: qmdresults
    class-message: qmdmessage
format: gfm
---

Hello.

What have I figured out about using Quarto/Mdsvex?:

1. `fig.path` needs to be set to `"./+page_files/"
    - you can then use the relativeImages npm package to properly link to output files

2. If you add **both** `fig.cap` and `fig.alt`, plot outputs will be wrapped with a `<figure>` tag. This could be helpful? Maybe the stuff can be styled now...

3. Adding `class-output` and `class-message` to `opts_chunk` wraps R code chunk output/messages with "``` qmdoutput". Not sure how this is then parsed by Mdsvex, but it's something?

4. Layouts ensure you only have to load CSS once?

## Todo

- [x] plot output (relative image paths)
- [ ] how to do math
- [x] code highlighting
    - gonna say this is done, 90% of what I've been looking for
- [ ] footnotes?
- [ ] citations?

## Styles?

Can we add styles?

Hell yes we can.

## Testing output/rendering

Here is a quote:

> The asphault city is my home. Furnished
> 
> From the outset with every sacramental prerequisite:
> 
> With newspapers. And tobacco. And brandy.
> 
> Distrustful and idle and contented to the end.

Here is some math.[^1]

$$
\int_0^1 x^3 + \theta x^2 dx
$$

Note that Math can also be inline, like this: $y = \beta_0 + \beta_1 x_1 + \epsilon$.

```{r}
library(tidyverse)

theme_set(
  theme_minimal(base_size = 15) +
    theme(
      panel.background = element_rect(fill = "transparent"),
      plot.background = element_rect(fill = "transparent")
    )
)
```

```{r}
summary(airquality)
```

Some more output.^[Pay this no mind.]

```{r}
lm(mpg ~ wt + cyl, data = mtcars) |> summary()
```

```{r}
mtcars |>
  group_by(vs) |>
  summarize(m = mean(mpg), s = sd(mpg)) |>
  knitr::kable(digits = 2)
```

Plots are messy though.^[Hey, you can also create in-line footnotes. They seem super convenient.]

```{r}
#| fig-align: center
#| fig-cap: "This is the caption."
#| fig-alt: "Maybe some alt-text?"
 
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) + geom_boxplot()
```

Sick, this is working now.[^2]

[^1]: There might be footnotes.

[^2]: You could consider writing something that's incredibly long as a footnote.

    These long footnotes could even span multiple paragraphs. However, note that new paragraphs need to be indented.
